---
title: "Utah Prison Population Projections"
subtitle: "Historical Trends and Population Projections"
format: 
  dashboard:
    orientation: rows  
    logo: images/DOCJ_WhitelogoMedium.svg
    theme: 
      - sandstone
      - custom-june4.scss
    #nav-buttons:
     # - linkedin
    scrolling: false
---

```{r}
#| label: setup
#| message: false
#| include: false


# Load required packages - streamlined to only what's needed for visualization
library(tidyverse)
library(plotly)
library(DT)
library(scales)
library(bslib)
library(bsicons)
library(lubridate)
library(kableExtra)
library(knitr)
library(readxl)
library(dplyr)
library(lubridate)
library(tsibble)
library(ggplot2)
library(plotly)
library(scales)
library(DT)
library(forecast)
library(htmltools)  # For tags$p function
library(fable)

# Set global options / Posit Connect Friendly options 
# Connect-friendly options
options(
  scipen = 999,
  warn = if(Sys.getenv("CONNECT_SERVER") != "") -1 else 0
)

theme_set(theme_minimal())

# Define parameters that would have been selected by the user
population_type <- "total"  # Options: "total", "male", "female"
projection_horizon <- 20    # Options: 8, 12, 20
scenario <- "baseline"      # Options: "baseline", "reform", "covid"

# Brand fonts for plotly
brand_title <- list(family = "UT Industry, serif", size = 20, color = "#071D49")
brand_body  <- list(family = "UT Industry, serif", size = 13, color = "#2C1E1B")

# === Capacity threshold brand colors (ADDED) =====================
cap_col_max   <- "#9B1B15"   # Absolute Maximum  (brand red)
cap_col_emerg <- "#A47217"   # Emergency         (brand gold)
cap_col_oper  <- "#2C1E1B"   # Operational       (brand brown)

# === Data currency stamp (ADDED) =================================
DATA_AS_OF <- "2024 Q4"
```

```{r}
#| label: load-preprocessed-data
#| include: false

# =================================================================
# FORECAST DATA — tidy long file (2026 vintage)
# date, quarter_label, year, quarter, series, kind,
# actual, central, low, high
#   series: Total | Male | Female
#   kind:   actual (2010 Q1–2026 Q2) | forecast (2026 Q3–2031 Q2)
# =================================================================

uppp <- read.csv("data/uppp_dashboard_quarterly.csv",
                 stringsAsFactors = FALSE) %>%
  mutate(
    date         = as.Date(date),
    Quarter_Year = yearquarter(quarter_label),
    across(c(actual, central, low, high), as.numeric)
  )

# --- Data currency, derived not hardcoded ----------------------

ADP_AS_OF <- uppp %>%
  filter(kind == "actual") %>%
  pull(quarter_label) %>%
  tail(1)                                  # "2026 Q2"

HORIZON_AS_OF <- uppp %>%
  filter(kind == "forecast") %>%
  pull(quarter_label) %>%
  tail(1)                                  # "2031 Q2"

DATA_AS_OF    <- ADP_AS_OF
BASE_LABEL    <- ADP_AS_OF
HORIZON_LABEL <- HORIZON_AS_OF
HORIZON_YEARS <- 5


# =================================================================
# SUMMARY METRICS
#
# Growth rates are READ from the published summary, not recomputed.
# A naive CAGR anchored on the last actual quarter understates
# female growth by ~0.4 pts, because female 2026 Q2 (462.3) sits
# about 8 above the prior five quarters (442–454).
#
# pct_change is derived FROM the published annual rate so the two
# value boxes stay internally consistent.
# =================================================================

uppp_summary    <- read.csv("data/uppp_dashboard_summary.csv",
                            stringsAsFactors = FALSE)
summary_central <- uppp_summary %>% filter(scenario == "central")

ann_growth <- list(
  Total  = summary_central$total_ann_growth,     # 2.11
  Male   = summary_central$male_ann_growth,      # 2.16
  Female = summary_central$female_ann_growth     # 1.36
)

# =================================================================
# CAPACITY THRESHOLDS
# Male/female are hardcoded in their own chunks; total is the sum.
# VERIFY these against the UDOC capacity source before publishing.
# =================================================================

capacity_thresholds <- list(
  Total  = list(absolute_max = 7326, emergency = 7180, operational = 7070),
  Male   = list(absolute_max = 6830, emergency = 6693.4, operational = 6590.95),
  Female = list(absolute_max =  496, emergency =  486.08, operational =  478.64)
)

series_meta <- list(
  Total  = list(color = "#007000", title = "Total Prison Population"),
  Male   = list(color = "#093692", title = "Male Prison Population"),
  Female = list(color = "#9B1B15", title = "Female Prison Population")
)

# =================================================================
# Assemble the object the Projections and Capacity chunks expect.
# Column names (Actual / Forecasted / Lower_CI / Upper_CI /
# Forecasted_hybrid / Percent_Change) match the pre-2026 vintage so
# those chunks run unchanged. Quarter_Year stays a yearquarter —
# the chunks rely on `max_date + 1`, `seq(by = 1)` and the "%q"
# format code, all of which are tsibble behaviours.
# =================================================================

build_population_data <- function(s) {

  fd <- uppp %>%
    filter(series == s) %>%
    arrange(date) %>%
    transmute(
      Quarter_Year = yearquarter(quarter_label),
      Actual       = if_else(kind == "actual",   actual,  NA_real_),
      Forecasted   = if_else(kind == "forecast", central, NA_real_),
      Lower_CI     = if_else(kind == "forecast", low,     NA_real_),
      Upper_CI     = if_else(kind == "forecast", high,    NA_real_)
    )

  td <- uppp %>%
    filter(series == s, kind == "forecast") %>%
    arrange(date) %>%
    transmute(
      Quarter_Year      = yearquarter(quarter_label),
      Forecasted_hybrid = central,
      Percent_Change    = round(100 * (central / lag(central) - 1), 1)
    )

  cur <- uppp %>% filter(series == s, kind == "actual")   %>%
                  pull(actual)  %>% tail(1)
  prj <- uppp %>% filter(series == s, kind == "forecast") %>%
                  pull(central) %>% tail(1)
  ann <- ann_growth[[s]]
  cap <- capacity_thresholds[[s]]

  list(
    forecast_data = fd,
    table_data    = td,
    color         = series_meta[[s]]$color,
    title         = series_meta[[s]]$title,

    current_pop   = cur,
    projected_pop = prj,
    annual_change = ann,
    pct_change    = round(100 * ((1 + ann / 100)^HORIZON_YEARS - 1), 1),

    capacity_absolute_max = cap$absolute_max,
    capacity_emergency    = cap$emergency,
    capacity_operational  = cap$operational
  )
}

metrics_total  <- build_population_data("Total")
metrics_male   <- build_population_data("Male")
metrics_female <- build_population_data("Female")

current_total  <- metrics_total$current_pop
current_male   <- metrics_male$current_pop
current_female <- metrics_female$current_pop

projected_total  <- metrics_total$projected_pop
pct_change_total <- metrics_total$pct_change

get_data_for_population <- function(pop) {
  switch(tolower(gsub(" ?Population", "", pop)),
         "male"   = metrics_male,
         "female" = metrics_female,
         metrics_total)
}


# --- Scenario bracket (for Projections page, if surfaced) ------

scenario_bracket <- uppp_summary %>%
  transmute(
    Scenario = recode(scenario,
                      central   = "Central (20/80)",
                      bracket_a = "Lower (0/100)",
                      bracket_b = "Upper (50/50)"),
    Weight            = weight,
    `2031 Q2 ADP`     = total_adp_5yr,
    `Annual growth %` = total_ann_growth
  )

# =================================================================
# DRIVER DATA — from DB_Input_v1_2026_Drivers_Plots
# =================================================================

adp_offense_type <- read.csv("data/Drivers/adp_offense_type_2026.csv") %>%
  mutate(Quarter_Year = yearquarter(Quarter_Year))

adp_offense_severity <- read.csv("data/Drivers/adp_offense_severity_2026.csv") %>%
  mutate(Quarter_Year = yearquarter(Quarter_Year))

adp_sex <- read.csv("data/Drivers/adp_sex_2026.csv") %>%
  mutate(Quarter_Year = yearquarter(Quarter_Year))

flows_quarterly <- read.csv("data/Drivers/flows_quarterly_2026.csv") %>%
  mutate(Quarter_Year = yearquarter(quarter)) %>% select(-quarter)

flows_by_type <- read.csv("data/Drivers/flows_by_offense_type_2026.csv") %>%
  mutate(Quarter_Year = yearquarter(quarter)) %>% select(-quarter)

flows_by_degree <- read.csv("data/Drivers/flows_by_offense_degree_2026.csv") %>%
  mutate(Quarter_Year = yearquarter(quarter)) %>% select(-quarter)

flows_by_type_f <- read.csv("data/Drivers/flows_by_offense_type_female_2026.csv") %>%
  mutate(Quarter_Year = yearquarter(quarter)) %>% select(-quarter)

FLOW_END <- max(flows_quarterly$Quarter_Year)
```


#  {.sidebar}

**DRAFT VERSION**

### About
<span style="color: #071D49">The Utah Prison Projection Project is a collaborative effort between the Utah Department of Corrections and the Utah Department of Criminal Justice. This initiative aims to develop a Utah-specific prison population forecasting tool to support criminal justice resource planning and analyze the impact of new legislation and policy on prison populations. </span>

<span style="color: #565656">**This dashboard provides projections of Utah's prison population based on historical data.**</span>

<span style="color: #565656">**Population Type:** Total, Male, Female </span>

<span style="color: #565656">**Projection Horizon:** 5-years </span>

<span style="color: #565656">**Capacity Values (for 2026):** Absolute Maximum Capacity (100% threshold), Emergency Capacity (98% threshold), Operational Capacity (96.5% threshold). *Note that capacity numbers are updated as of 2026 Q2 and fluctuate through time.</span>*



<!-- ADDED: plain-language takeaway + data currency stamp -->
<span style="color: #071D49"><b>At a glance:</b> Utah's total prison population is projected to grow `r pct_change_total`% over five years &mdash; from `r comma(round(current_total))` in `r BASE_LABEL` to about `r comma(round(projected_total))` by `r HORIZON_LABEL`. </span>

<span style="color: #565656"><i>Data current as of `r DATA_AS_OF`.</i></span>



# 🏛️ Overview {orientation="row"}

## Value Boxes {height="30%"}

```{r}
#| label: value-boxes

# Get data based on the selected population type
data <- get_data_for_population(population_type)

# Create layout with value boxes
layout_column_wrap(
  width = 1/4,  # exactly 4 columns; shrink together
  fixed_width = FALSE,
  heights_equal = "row",
  gap = "0.75rem",
  value_box(
    title = paste(str_to_title(gsub("Population", "", population_type)), "Current Population"),
    value = comma(round(data$current_pop)),
    showcase = bs_icon("people-fill"),
    showcase_layout = showcase_left_center(),
    theme = value_box_theme(bg = "#A6C4CA", fg = "#2C1E1B"),
    tags$p(BASE_LABEL)                                       # was "2024 Q4"
  ),
  value_box(
    title = paste(str_to_title(gsub("Population", "", population_type)), "Projected Population"),
    value = comma(round(data$projected_pop)),
    showcase = bs_icon("graph-up"),
    showcase_layout = showcase_left_center(),
    theme = value_box_theme(bg = "#007000", fg = "#FFFFFF"),
    tags$p(paste0("5-year projection · ", HORIZON_LABEL))   # replaces both tags$p lines
  ),
  value_box(
    title = "Percent Change",
    value = paste0(data$pct_change, "%"),
    showcase = bs_icon("percent"),
    showcase_layout = showcase_left_center(),
    theme = value_box_theme(bg = "#A47217", fg = "#FFFFFF"),
    tags$p("over 5 years")
  ),
  value_box(
    title = "Average Annual Change",
    value = paste0(data$annual_change, "%"),
    showcase = bs_icon("calendar-event"),
    showcase_layout = showcase_left_center(),
    theme = value_box_theme(bg = "#9B1B15", fg = "#FFFFFF"),
    tags$p("annually")
  )
)
```

## Plot: Quarterly Net Difference Prison Admissions and Releases Over Time {height="60%"}

```{r}
#| label: admissions-releases-plot
#| title: Prison Admissions & Releases (Net Difference)
#| height: 60%

# Look up the line's value at each annotated quarter so the arrowhead
# lands on the series rather than floating in empty space.
val_at <- function(q) {
  v <- flows_quarterly$net_admissions[
    flows_quarterly$Quarter_Year == yearquarter(q)]
  if (length(v) == 0) 0 else v[1]
}

y_jri   <- val_at("2015 Q1")
y_covid <- val_at("2020 Q1")

plot_ly(flows_quarterly, x = ~as.character(Quarter_Year)) %>%
  add_lines(
    y = ~net_admissions,
    name = "Net difference",
    line = list(color = "#093692", width = 2.5)          # brand navy
  ) %>%
  add_lines(
    y = 0,
    name = "No growth line",
    line = list(color = "#A47217", dash = "dash", width = 2)   # brand gold
  ) %>%
  layout(
    title = "",
    font = brand_body,                      # UT Industry, #2C1E1B
    xaxis = list(
      title = "Year-Quarter",
      type = "category",
      dtick = 4,
      zerolinecolor = "#A47217",
      zerolinewidth = 2
    ),
    yaxis = list(
      title = "Count",
      zerolinecolor = "#A47217",
      zerolinewidth = 2
    ),
    plot_bgcolor = "#f5f5f5",
    paper_bgcolor = "#ffffff",
    legend = list(font = brand_body)
  ) %>%
  # Add annotations
  add_annotations(
    x = "2015 Q1",
    y = y_jri,
    text = "JRI",
    showarrow = TRUE,
    arrowhead = 2,
    arrowsize = 1,
    arrowwidth = 2,
    arrowcolor = "#2C1E1B",                 # brand brown
    ax = 0,
    ay = 40,   # label sits 40px below; arrow points up to the line
    font = list(family = "UT Industry, serif", size = 12, color = "#071D49"),
    bgcolor = "rgba(255, 255, 255, 0.85)",
    bordercolor = "#2C1E1B",
    borderwidth = 1
  ) %>%
  add_annotations(
    x = "2020 Q1",
    y = y_covid,
    text = "COVID",
    showarrow = TRUE,
    arrowhead = 2,
    arrowsize = 1,
    arrowwidth = 2,
    arrowcolor = "#2C1E1B",
    ax = 0,
    ay = 25,   # staggered vs. JRI so the labels don't collide near zero
    font = list(family = "UT Industry, serif", size = 12, color = "#071D49"),
    bgcolor = "rgba(255, 255, 255, 0.85)",
    bordercolor = "#2C1E1B",
    borderwidth = 1
  )
```

# 📈 Projections {.tabset}

## Total Population {height="275px"}

### Column {width=70%}

```{r}
#| label: total-projection-plot

# Get the appropriate forecast data for total population
total_data <- get_data_for_population("total")
forecast_data <- total_data$forecast_data
color_actual <- total_data$color
title <- paste(total_data$title, "Projection")

# Find the transition point between actual and forecast
actual_data <- forecast_data[!is.na(forecast_data$Actual),]
max_date <- max(actual_data$Quarter_Year)

# Get horizon dates for filtering
horizon_dates <- seq(max_date + 1, by = 1, length.out = projection_horizon)

# Create plot
p <- plot_ly(
  hoverlabel = list(namelength = -1)  # Don't truncate trace names in hover
)

# Add historical data
p <- p %>% add_trace(
  data = forecast_data[!is.na(forecast_data$Actual),],
  x = ~Quarter_Year,
  y = ~Actual,
  type = 'scatter',
  mode = 'lines',
  name = "Historical",
  line = list(color = "#007000"),
  hoverinfo = "text",
  text = ~paste0(
    format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
    "<br>Population: ", formatC(Actual, format="d", big.mark=",")
  )
)

# Add forecast data
forecast_part <- forecast_data[forecast_data$Quarter_Year %in% horizon_dates,]

# ADDED: 80% prediction-interval ribbon, drawn UNDER the projection line.
# Guarded so it silently no-ops if a forecast df lacks CI columns (won't break render).
#if (all(c("Lower_CI", "Upper_CI") %in% names(forecast_part))) {
 # p <- p %>% add_ribbons(
  #  data = forecast_part,
  #  x = ~Quarter_Year,
  # ymin = ~Lower_CI,
  #  ymax = ~Upper_CI,
  #  name = "Projection range",
  #  line = list(width = 0),
  #  fillcolor = "rgba(164, 114, 23, 0.18)",   # translucent brand gold
  #  hoverinfo = "skip"
  #)
#}

p <- p %>% add_trace(
  data = forecast_part,
  x = ~Quarter_Year,
  y = ~Forecasted,
  type = 'scatter',
  mode = 'lines',
  name = "Projection",
  line = list(color = "#A47217", dash = "dash"),
  hoverinfo = "text",
  text = ~paste0(
    format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
    "<br>Projected: ", formatC(Forecasted, format="d", big.mark=",")
  )
)

# Get all years for x-axis ticks
all_quarters <- unique(forecast_data$Quarter_Year)
all_years <- unique(format(all_quarters, "%Y"))

# Select every other year for x-axis ticks, using Q1 of each year
tick_positions <- c()
tick_labels <- c()

for (i in seq(1, length(all_years), by = 2)) {
  year <- all_years[i]
  year_quarters <- all_quarters[format(all_quarters, "%Y") == year]
  
  if (length(year_quarters) > 0) {
    # Find Q1 or closest quarter
    months <- as.numeric(format(year_quarters, "%m"))
    q1_idx <- which.min(abs(months - 2))  # closest to February (middle of Q1)
    q1_date <- year_quarters[q1_idx]
    
    tick_positions <- c(tick_positions, q1_date)
    tick_labels <- c(tick_labels, paste0(year, " Q", ceiling(months[q1_idx]/3)))
  }
}

# Complete layout
p %>% layout(
  title = list(
    text = title,
    font = list(family = "UT Industry, serif", size = 20, color = "#071D49"),
    y = 0.95,                          # nudge title down slightly from the very top
    yanchor = "top",
    pad = list(b = 10)                 # padding between title and plot
  ),
  margin = list(t = 70),               # <-- the key fix: room above the plot for the title
  font = list(family = "UT Industry, serif"),
  xaxis = list(
    title = "Year-Quarter",
    type = "date",
    tickmode = "array",
    tickvals = tick_positions,
    ticktext = tick_labels,
    tickangle = 0
  ),
  yaxis = list(
    title = "Prison Population",
    tickformat = ","
  ),
  hovermode = "closest",
  showlegend = TRUE
)


```

### Column {width=30%}

```{r}
#| label: total-projection-table

# Get the correct table data for total population
table_data <- total_data$table_data

# Limit rows to selected horizon
table_data <- head(table_data, projection_horizon)

# Format table
table_data <- table_data %>%
  mutate(
    Quarter_Year = as.character(Quarter_Year),
    Forecasted_hybrid = round(Forecasted_hybrid)
#     Lower_CI = round(Lower_CI),
#     Upper_CI = round(Upper_CI)
  ) %>%
  select(
    Quarter = Quarter_Year,
    Forecast = Forecasted_hybrid,
#     `Lower CI` = Lower_CI,
#     `Upper CI` = Upper_CI,
    `% Change` = Percent_Change
  )

datatable(
  table_data,
  extensions = "Buttons",                       # ADDED: download/copy buttons
  options = list(
    pageLength = projection_horizon,
    dom = 'Bt',                                 # B = buttons, t = table
    ordering = FALSE,
    buttons = list(
      list(extend = "csv",   text = "CSV"),
      list(extend = "excel", text = "Excel"),
      list(extend = "copy",  text = "Copy")
    )
  ),
  rownames = FALSE,
  caption = "Total Population Projections"
) %>%
  formatStyle(
    '% Change',
    color = styleInterval(c(0), c('#9B1B15', '#007000')),
    fontWeight = 'bold'
  )
```

## Male Population {height="275px"}

### Column {width=70%}

```{r}
#| label: male-projection-plot

# Get the appropriate forecast data for male population
male_data <- get_data_for_population("male")
forecast_data <- male_data$forecast_data
color_actual <- male_data$color
title <- paste(male_data$title, "Projection")

# Find the transition point between actual and forecast
actual_data <- forecast_data[!is.na(forecast_data$Actual),]
max_date <- max(actual_data$Quarter_Year)

# Get horizon dates for filtering
horizon_dates <- seq(max_date + 1, by = 1, length.out = projection_horizon)

# Create plot
p <- plot_ly(
  hoverlabel = list(namelength = -1)  # Don't truncate trace names in hover
)

# Add historical data
p <- p %>% add_trace(
  data = forecast_data[!is.na(forecast_data$Actual),],
  x = ~Quarter_Year,
  y = ~Actual,
  type = 'scatter',
  mode = 'lines',
  name = "Historical",
  line = list(color = color_actual),
  hoverinfo = "text",
  text = ~paste0(
    format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
    "<br>Population: ", formatC(Actual, format="d", big.mark=",")
  )
)

# Add forecast data
forecast_part <- forecast_data[forecast_data$Quarter_Year %in% horizon_dates,]

# ADDED: 80% prediction-interval ribbon, drawn UNDER the projection line.
# Guarded so it silently no-ops if a forecast df lacks CI columns (won't break render).
#if (all(c("Lower_CI", "Upper_CI") %in% names(forecast_part))) {
#  p <- p %>% add_ribbons(
#    data = forecast_part,
#    x = ~Quarter_Year,
#    ymin = ~Lower_CI,
#    ymax = ~Upper_CI,
#    name = "Projection range",
#    line = list(width = 0),
#    fillcolor = "rgba(164, 114, 23, 0.18)",   # translucent brand gold
#    hoverinfo = "skip"
#  )
#}

p <- p %>% add_trace(
  data = forecast_part,
  x = ~Quarter_Year,
  y = ~Forecasted,
  type = 'scatter',
  mode = 'lines',
  name = "Projection",
  line = list(color = "#A47217", dash = "dash"),
  hoverinfo = "text",
  text = ~paste0(
    format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
    "<br>Projected: ", formatC(Forecasted, format="d", big.mark=",")
  )
)

# Get all years for x-axis ticks
all_quarters <- unique(forecast_data$Quarter_Year)
all_years <- unique(format(all_quarters, "%Y"))

# Select every other year for x-axis ticks, using Q1 of each year
tick_positions <- c()
tick_labels <- c()

for (i in seq(1, length(all_years), by = 2)) {
  year <- all_years[i]
  year_quarters <- all_quarters[format(all_quarters, "%Y") == year]
  
  if (length(year_quarters) > 0) {
    # Find Q1 or closest quarter
    months <- as.numeric(format(year_quarters, "%m"))
    q1_idx <- which.min(abs(months - 2))  # closest to February (middle of Q1)
    q1_date <- year_quarters[q1_idx]
    
    tick_positions <- c(tick_positions, q1_date)
    tick_labels <- c(tick_labels, paste0(year, " Q", ceiling(months[q1_idx]/3)))
  }
}

# Complete layout
p %>% layout(
  title = list(
    text = title,
    font = list(family = "UT Industry, serif", size = 20, color = "#071D49"),
    y = 0.95,                          # nudge title down slightly from the very top
    yanchor = "top",
    pad = list(b = 10)                 # padding between title and plot
  ),
  margin = list(t = 70),               # <-- the key fix: room above the plot for the title
  font = list(family = "UT Industry, serif"),
  xaxis = list(
    title = "Year-Quarter",
    type = "date",
    tickmode = "array",
    tickvals = tick_positions,
    ticktext = tick_labels,
    tickangle = 0
  ),
  yaxis = list(
    title = "Prison Population",
    tickformat = ","
  ),
  hovermode = "closest",
  showlegend = TRUE
)

```

### Column {width=30%}

```{r}
#| label: male-projection-table

# Get the correct table data for male population
table_data <- male_data$table_data

# Limit rows to selected horizon
table_data <- head(table_data, projection_horizon)

# Format table
table_data <- table_data %>%
  mutate(
    Quarter_Year = as.character(Quarter_Year),
    Forecasted_hybrid = round(Forecasted_hybrid)
#     Lower_CI = round(Lower_CI),
#     Upper_CI = round(Upper_CI)
  ) %>%
  select(
    Quarter = Quarter_Year,
    Forecast = Forecasted_hybrid,
#     `Lower CI` = Lower_CI,
#     `Upper CI` = Upper_CI,
    `% Change` = Percent_Change
  )

datatable(
  table_data,
  extensions = "Buttons",                       # ADDED: download/copy buttons
  options = list(
    pageLength = projection_horizon,
    dom = 'Bt',                                 # B = buttons, t = table
    ordering = FALSE,
    buttons = list(
      list(extend = "csv",   text = "CSV"),
      list(extend = "excel", text = "Excel"),
      list(extend = "copy",  text = "Copy")
    )
  ),
  rownames = FALSE,
  caption = "Male Population Projections"
) %>%
  formatStyle(
    '% Change',
    color = styleInterval(c(0), c('#9B1B15', '#007000')),
    fontWeight = 'bold'
  )
```

## Female Population {height="275px"}

### Column {width=70%}

```{r}
#| label: female-projection-plot

# Get the appropriate forecast data for female population
female_data <- get_data_for_population("female")
forecast_data <- female_data$forecast_data
color_actual <- female_data$color
title <- paste(female_data$title, "Projection")

# Find the transition point between actual and forecast
actual_data <- forecast_data[!is.na(forecast_data$Actual),]
max_date <- max(actual_data$Quarter_Year)

# Get horizon dates for filtering
horizon_dates <- seq(max_date + 1, by = 1, length.out = projection_horizon)

# Create plot
p <- plot_ly(
  hoverlabel = list(namelength = -1)  # Don't truncate trace names in hover
)

# Add historical data
p <- p %>% add_trace(
  data = forecast_data[!is.na(forecast_data$Actual),],
  x = ~Quarter_Year,
  y = ~Actual,
  type = 'scatter',
  mode = 'lines',
  name = "Historical",
  line = list(color = color_actual),
  hoverinfo = "text",
  text = ~paste0(
    format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
    "<br>Population: ", formatC(Actual, format="d", big.mark=",")
  )
)

# Add forecast data
forecast_part <- forecast_data[forecast_data$Quarter_Year %in% horizon_dates,]

# ADDED: 80% prediction-interval ribbon, drawn UNDER the projection line.
# Guarded so it silently no-ops if a forecast df lacks CI columns (won't break render).
#if (all(c("Lower_CI", "Upper_CI") %in% names(forecast_part))) {
#  p <- p %>% add_ribbons(
#    data = forecast_part,
#    x = ~Quarter_Year,
#    ymin = ~Lower_CI,
#    ymax = ~Upper_CI,
#    name = "Projection range",
#    line = list(width = 0),
#    fillcolor = "rgba(164, 114, 23, 0.18)",   # translucent brand gold
#    hoverinfo = "skip"
#  )
#}

p <- p %>% add_trace(
  data = forecast_part,
  x = ~Quarter_Year,
  y = ~Forecasted,
  type = 'scatter',
  mode = 'lines',
  name = "Projection",
  line = list(color = "#A47217", dash = "dash"),
  hoverinfo = "text",
  text = ~paste0(
    format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
    "<br>Projected: ", formatC(Forecasted, format="d", big.mark=",")
  )
)

# Get all years for x-axis ticks
all_quarters <- unique(forecast_data$Quarter_Year)
all_years <- unique(format(all_quarters, "%Y"))

# Select every other year for x-axis ticks, using Q1 of each year
tick_positions <- c()
tick_labels <- c()

for (i in seq(1, length(all_years), by = 2)) {
  year <- all_years[i]
  year_quarters <- all_quarters[format(all_quarters, "%Y") == year]
  
  if (length(year_quarters) > 0) {
    # Find Q1 or closest quarter
    months <- as.numeric(format(year_quarters, "%m"))
    q1_idx <- which.min(abs(months - 2))  # closest to February (middle of Q1)
    q1_date <- year_quarters[q1_idx]
    
    tick_positions <- c(tick_positions, q1_date)
    tick_labels <- c(tick_labels, paste0(year, " Q", ceiling(months[q1_idx]/3)))
  }
}

# Complete layout
p %>% layout(
  title = list(
    text = title,
    font = list(family = "UT Industry, serif", size = 20, color = "#071D49"),
    y = 0.95,                          # nudge title down slightly from the very top
    yanchor = "top",
    pad = list(b = 10)                 # padding between title and plot
  ),
  margin = list(t = 70),               # <-- the key fix: room above the plot for the title
  font = list(family = "UT Industry, serif"),
  xaxis = list(
    title = "Year-Quarter",
    type = "date",
    tickmode = "array",
    tickvals = tick_positions,
    ticktext = tick_labels,
    tickangle = 0
  ),
  yaxis = list(
    title = "Prison Population",
    tickformat = ","
  ),
  hovermode = "closest",
  showlegend = TRUE
)
```

### Column {width=30%}

```{r}
#| label: female-projection-table

# Get the correct table data for female population
table_data <- female_data$table_data

# Limit rows to selected horizon
table_data <- head(table_data, projection_horizon)

# Format table
table_data <- table_data %>%
  mutate(
    Quarter_Year = as.character(Quarter_Year),
    Forecasted_hybrid = round(Forecasted_hybrid)
#    Lower_CI = round(Lower_CI),
#    Upper_CI = round(Upper_CI)
  ) %>%
  select(
    Quarter = Quarter_Year,
    Forecast = Forecasted_hybrid,
#     `Lower CI` = Lower_CI,
#     `Upper CI` = Upper_CI,
    `% Change` = Percent_Change
  )

datatable(
  table_data,
  extensions = "Buttons",                       # ADDED: download/copy buttons
  options = list(
    pageLength = projection_horizon,
    dom = 'Bt',                                 # B = buttons, t = table
    ordering = FALSE,
    buttons = list(
      list(extend = "csv",   text = "CSV"),
      list(extend = "excel", text = "Excel"),
      list(extend = "copy",  text = "Copy")
    )
  ),
  rownames = FALSE,
  caption = "Female Population Projections"
) %>%
  formatStyle(
    '% Change',
    color = styleInterval(c(0), c('#9B1B15', '#007000')),  # brand red below 0, brand green at/above 0
    fontWeight = 'bold'
  )
```



# ↕️️ Trends

::: panel-tabset
### Primary Offense Type
```{r}
#| label: trend-offense-type
#| height: 800

# Collapse the two sex-offense categories into one panel. ADP is a
# count, so summing across the two categories within each quarter is
# the correct aggregation.
adp_type_combined <- adp_offense_type %>%
  mutate(Category = if_else(
    Category %in% c("SEX/REGISTERABLE", "SEX/NON-REGISTERABLE"),
    "SEX (REGISTERABLE & NON-REGISTERABLE)",
    Category
  )) %>%
  group_by(Quarter_Year, Category) %>%
  summarize(Count = sum(Count, na.rm = TRUE), .groups = "drop")


p <- ggplot(
  adp_type_combined,
  aes(
    x = Quarter_Year,
    y = Count,
    group = 1,
    text = paste(
      "Offense Type:", Category,
      "<br>Quarter:", Quarter_Year,
      "<br>ADP:", round(Count, 0)
    )
  )
) +
  geom_line(
    linewidth = 1,
    color = "#093692"
  ) +
  geom_vline(
    xintercept = yearquarter("2020 Q1"),
    linetype = "dashed",
    color = "#C6D0BE"
  ) +
  facet_wrap(
    ~Category,
    ncol = 3,
    scales = "free_y"
  ) +
  expand_limits(y = 0) +
  labs(
    title = "Utah Prison Population by Primary Offense Type",
    x = "Year Quarter",
    y = "Average Daily Population"
  ) +
 theme_minimal() +
  theme(
    plot.title    = element_text(face = "bold", size = 16, color = "#071D49",
                                 margin = margin(b = 15)),
    axis.title.x  = element_text(size = 11, color = "#2C1E1B",
                                 margin = margin(t = 20)),   # space above "Year Quarter"
    axis.title.y  = element_text(size = 11, color = "#2C1E1B",
                                 margin = margin(r = 15)),
    strip.text    = element_text(face = "bold", size = 9, color = "#071D49",
                                 margin = margin(b = 8)),    # space below panel titles
    axis.text.x   = element_text(angle = 45, hjust = 1, size = 8, color = "#2C1E1B",
                                 margin = margin(t = 5)),
    axis.text.y   = element_text(size = 8, color = "#2C1E1B"),
    panel.spacing.x = unit(1.5, "lines"),   # horizontal gap between panels
    panel.spacing.y = unit(2, "lines"),     # vertical gap between panel rows
    plot.margin   = margin(t = 10, r = 15, b = 20, l = 10)
  )

ggplotly(p, tooltip = "text") %>%
  layout(
    font = list(family = "UT Industry, serif"),
    margin = list(b = 90, t = 60, l = 70, r = 30)
  )
```

### Offense Severity
```{r}
#| label: trend-offense-severity
#| height: 700

# Felony degrees only. MA, S, and "No Active Offense" are near-zero
# flat lines that add panels without signal. MB excluded upstream.
sev_keep <- adp_offense_severity %>%
  filter(!Category %in% c("MA", "S", "No Active Offense")) %>%
  mutate(Category = recode(Category,
    "F1" = "F1 — First degree felony",
    "F2" = "F2 — Second degree felony",
    "F3" = "F3 — Third degree felony",
    "FC" = "FC — Capital felony"
  ))

p <- ggplot(
  sev_keep,
  aes(
    x = Quarter_Year,
    y = Count,
    group = 1,
    text = paste(
      "Severity:", Category,
      "<br>Quarter:", Quarter_Year,
      "<br>ADP:", round(Count, 0)
    )
  )
) +
  geom_line(
    linewidth = 1,
    color = "#007000"
  ) +
  geom_vline(
    xintercept = yearquarter("2020 Q1"),
    linetype = "dashed",
    color = "#C6D0BE"
  ) +
  facet_wrap(
    ~Category,
    ncol = 2
  ) +
  labs(
    title = "Utah Prison Population by Offense Severity",
    x = "Year Quarter",
    y = "Average Daily Population"
  ) +
theme_minimal() +
  theme(
    plot.title    = element_text(face = "bold", size = 16, color = "#071D49",
                                 margin = margin(b = 15)),
    axis.title.x  = element_text(size = 11, color = "#2C1E1B",
                                 margin = margin(t = 20)),   # space above "Year Quarter"
    axis.title.y  = element_text(size = 11, color = "#2C1E1B",
                                 margin = margin(r = 15)),
    strip.text    = element_text(face = "bold", size = 9, color = "#071D49",
                                 margin = margin(b = 8)),    # space below panel titles
    axis.text.x   = element_text(angle = 45, hjust = 1, size = 8, color = "#2C1E1B",
                                 margin = margin(t = 5)),
    axis.text.y   = element_text(size = 8, color = "#2C1E1B"),
    panel.spacing.x = unit(1.5, "lines"),   # horizontal gap between panels
    panel.spacing.y = unit(2, "lines"),     # vertical gap between panel rows
    plot.margin   = margin(t = 10, r = 15, b = 20, l = 10)
  )

ggplotly(p, tooltip = "text") %>%
  layout(
    font = list(family = "UT Industry, serif"),
    margin = list(b = 90, t = 60, l = 70, r = 30)
  )
```

### Net Admissions by Offense Degree
```{r}
#| label: trend-net-adm-degree
#| height: 600

# Full labels on the strips, so the codes are defined on the panel
# itself rather than in a legend. (A color legend can't be used here —
# mapping color to the facet variable breaks ggplotly's panel/trace
# assignment.)
deg_labeled <- flows_by_degree %>%
  filter(!primary_ofnse_degree %in% c("FC")) %>%
  mutate(primary_ofnse_degree = recode(primary_ofnse_degree,
    "F1" = "F1 — First degree felony",
    "F2" = "F2 — Second degree felony",
    "F3" = "F3 — Third degree felony",
    #"FC" = "FC — Capital felony"
  ))

p <- ggplot(
  deg_labeled,
  aes(
    x = Quarter_Year,
    y = net_admissions,
    group = 1,
    text = paste(
      "Offense Degree:", primary_ofnse_degree,
      "<br>Quarter:", Quarter_Year,
      "<br>Net Admissions:", net_admissions,
      "<br>Admissions:", admissions,
      "<br>Releases:", releases
    )
  )
) +
  geom_line(
    linewidth = 1,
    color = "#007000"
  ) +
  geom_hline(
    yintercept = 0,
    color = "#A47217",
    linetype = "dashed"
  ) +
  facet_wrap(
    ~primary_ofnse_degree,
    ncol = 2,
    scales = "free_y"
  ) +
  labs(
    title = "Net Admissions by Primary Offense Degree",
    x = "Year Quarter",
    y = "Net Admissions"
  ) +
  theme_minimal() +
  theme(
    plot.title    = element_text(face = "bold", size = 16, color = "#071D49",
                                 margin = margin(b = 15)),
    axis.title.x  = element_text(size = 11, color = "#2C1E1B",
                                 margin = margin(t = 20)),
    axis.title.y  = element_text(size = 11, color = "#2C1E1B",
                                 margin = margin(r = 15)),
    strip.text    = element_text(face = "bold", size = 9, color = "#071D49",
                                 margin = margin(b = 8)),
    axis.text.x   = element_text(angle = 45, hjust = 1, size = 8, color = "#2C1E1B",
                                 margin = margin(t = 5)),
    axis.text.y   = element_text(size = 8, color = "#2C1E1B"),
    panel.spacing.x = unit(1.5, "lines"),
    panel.spacing.y = unit(2, "lines"),
    plot.margin   = margin(t = 10, r = 15, b = 20, l = 10)
  )

ggplotly(p, tooltip = "text") %>%
  layout(
    font = list(family = "UT Industry, serif"),
    margin = list(b = 90, t = 60, l = 70, r = 30)
  )
```
::: 


# ⚠️ Capacity {.tabset}

## Total Population {height="420px"}

```{r}
#| label: total-capacity-analysis

# Get the appropriate forecast data for total population
total_data <- get_data_for_population("total")
forecast_data <- total_data$forecast_data
color_actual <- total_data$color
title <- paste(total_data$title, "vs. Capacity")

# Get the capacity values directly from the function return values
absolute_max_capacity <- total_data$capacity_absolute_max
emergency_capacity <- total_data$capacity_emergency
operational_capacity <- total_data$capacity_operational

# Find the transition point between actual and forecast
actual_data <- forecast_data[!is.na(forecast_data$Actual),]
max_date <- max(actual_data$Quarter_Year)

# Get horizon dates for filtering
horizon_dates <- seq(max_date + 1, by = 1, length.out = projection_horizon)

# Create capacity dataframe with all three capacity types
capacity_df <- data.frame(
  Quarter_Year = rep(unique(forecast_data$Quarter_Year), 3),
  Capacity = c(
    rep(absolute_max_capacity, length(unique(forecast_data$Quarter_Year))),
    rep(emergency_capacity, length(unique(forecast_data$Quarter_Year))),
    rep(operational_capacity, length(unique(forecast_data$Quarter_Year)))
  ),
  Type = c(
    rep("Absolute Maximum Capacity", length(unique(forecast_data$Quarter_Year))),
    rep("Emergency Capacity", length(unique(forecast_data$Quarter_Year))),
    rep("Operational Capacity", length(unique(forecast_data$Quarter_Year)))
  )
)

# Calculate when each capacity is reached
capacity_reached <- list(
  absolute_max = NA,
  emergency = NA,
  operational = NA
)

for (i in 1:length(horizon_dates)) {
  date <- horizon_dates[i]
  projected_value <- forecast_data$Forecasted[forecast_data$Quarter_Year == date]

  if (!is.na(projected_value)) {
    if (is.na(capacity_reached$absolute_max) && projected_value > absolute_max_capacity) {
      capacity_reached$absolute_max <- date
    }
    if (is.na(capacity_reached$emergency) && projected_value > emergency_capacity) {
      capacity_reached$emergency <- date
    }
    if (is.na(capacity_reached$operational) && projected_value > operational_capacity) {
      capacity_reached$operational <- date
    }
  }
}

# Create plot
p <- plot_ly(
  hoverlabel = list(namelength = -1)
) %>%
  add_trace(
    data = forecast_data[!is.na(forecast_data$Actual),],
    x = ~Quarter_Year,
    y = ~Actual,
    type = 'scatter',
    mode = 'lines',
    name = "Historical",
    line = list(color = "#007000"),
    hoverinfo = "text",
    text = ~paste0(
      format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
      "<br>Population: ", formatC(Actual, format="d", big.mark=",")
    )
  )

forecast_part <- forecast_data[forecast_data$Quarter_Year %in% horizon_dates,]
p <- p %>%
  add_trace(
    data = forecast_part,
    x = ~Quarter_Year,
    y = ~Forecasted,
    type = 'scatter',
    mode = 'lines',
    name = "Projection",
    line = list(color = "#A47217", dash = "dash"),
    hoverinfo = "text",
    text = ~paste0(
      format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
      "<br>Projected: ", formatC(Forecasted, format="d", big.mark=",")
    )
  )

# Capacity threshold lines
p <- p %>%
  add_trace(
    data = capacity_df[capacity_df$Type == "Absolute Maximum Capacity",],
    x = ~Quarter_Year, y = ~Capacity,
    type = 'scatter', mode = 'lines',
    name = "Absolute Maximum Capacity",
    line = list(color = cap_col_max, dash = "dot", width = 2),
    hoverinfo = "text",
    text = ~paste0(
      format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
      "<br>Absolute Maximum Capacity: ", formatC(Capacity, format="d", big.mark=",")
    )
  ) %>%
  add_trace(
    data = capacity_df[capacity_df$Type == "Emergency Capacity",],
    x = ~Quarter_Year, y = ~Capacity,
    type = 'scatter', mode = 'lines',
    name = "Emergency Capacity",
    line = list(color = cap_col_emerg, dash = "dot", width = 2),
    hoverinfo = "text",
    text = ~paste0(
      format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
      "<br>Emergency Capacity: ", formatC(Capacity, format="d", big.mark=",")
    )
  ) %>%
  add_trace(
    data = capacity_df[capacity_df$Type == "Operational Capacity",],
    x = ~Quarter_Year, y = ~Capacity,
    type = 'scatter', mode = 'lines',
    name = "Operational Capacity",
    line = list(color = cap_col_oper, dash = "dot", width = 2),
    hoverinfo = "text",
    text = ~paste0(
      format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
      "<br>Operational Capacity: ", formatC(Capacity, format="d", big.mark=",")
    )
  )

# Threshold-crossing annotations, staggered on both axes so the three
# arrows (which converge within ~2 years) don't overlap.
for (capacity_type in names(capacity_reached)) {
  date <- capacity_reached[[capacity_type]]
  if (!is.na(date)) {
    projected_value <- forecast_data$Forecasted[forecast_data$Quarter_Year == date]

    year <- format(date, "%Y")
    quarter <- paste0("Q", ceiling(as.numeric(format(date, "%m"))/3))
    quarter_label <- paste0(year, " ", quarter)

    capacity_name <- switch(capacity_type,
                            "absolute_max" = "Absolute Maximum Capacity",
                            "emergency"    = "Emergency Capacity",
                            "operational"  = "Operational Capacity")

    annotation_color <- switch(capacity_type,
                               "absolute_max" = cap_col_max,
                               "emergency"    = cap_col_emerg,
                               "operational"  = cap_col_oper)

    p <- p %>%
      add_annotations(
        x = date,
        y = projected_value,
        text = paste(quarter_label),
        showarrow = TRUE,
        arrowhead = 2,
        arrowsize = 1,
        arrowwidth = 1.5,
        arrowcolor = annotation_color,
        ax = switch(capacity_type,
                    "absolute_max" =  70,
                    "emergency"    = -80,
                    "operational"  = -80),
        ay = switch(capacity_type,
                    "absolute_max" = -70,
                    "emergency"    = -35,
                    "operational"  =  45),
        font = list(color = annotation_color, size = 11,
                    family = "UT Industry, serif"),
        bgcolor = "rgba(255, 255, 255, 0.88)",
        bordercolor = annotation_color,
        borderwidth = 1,
        borderpad = 4,
        align = "center"
      )
  }
}

all_quarters <- unique(forecast_data$Quarter_Year)

y_top <- max(c(absolute_max_capacity, forecast_data$Forecasted, forecast_data$Actual),
             na.rm = TRUE) * 1.03

p %>% layout(
  title = list(
    text = title,
    font = list(family = "UT Industry, serif", size = 18, color = "#071D49"),
    y = 0.96,
    yanchor = "top"
  ),
  margin = list(t = 60, b = 50, l = 70, r = 30),
  font = list(family = "UT Industry, serif", size = 11),
  showlegend = TRUE,
  shapes = list(list(
    type = "rect", xref = "paper", yref = "y",
    x0 = 0, x1 = 1, y0 = operational_capacity, y1 = y_top,
    fillcolor = "rgba(155, 27, 21, 0.07)",
    line = list(width = 0), layer = "below"
  )),
  xaxis = list(
    title = "Year-Quarter",
    type = "date",
    tickmode = "array",
    tickvals = all_quarters[seq(1, length(all_quarters), by = 8)],
    ticktext = format(all_quarters[seq(1, length(all_quarters), by = 8)], "%Y"),
    tickangle = 0
  ),
  yaxis = list(
    title = "Prison Population",
    tickformat = ",",
    range = c(5000, 7600)      # shared with the male chart for comparability
  ),
  hovermode = "closest"
)
```

## Male Population {height="420px"}

```{r}
#| label: male-capacity-analysis

male_data <- get_data_for_population("male")
forecast_data <- male_data$forecast_data
color_actual <- male_data$color
title <- paste(male_data$title, "vs. Capacity")

absolute_max_capacity <- 6588  # Male-specific absolute maximum capacity
emergency_capacity <- 6456     # VERIFY against UDC source
operational_capacity <- 6357   # VERIFY against UDC source

actual_data <- forecast_data[!is.na(forecast_data$Actual),]
max_date <- max(actual_data$Quarter_Year)

horizon_dates <- seq(max_date + 1, by = 1, length.out = projection_horizon)

capacity_df <- data.frame(
  Quarter_Year = rep(unique(forecast_data$Quarter_Year), 3),
  Capacity = c(
    rep(absolute_max_capacity, length(unique(forecast_data$Quarter_Year))),
    rep(emergency_capacity, length(unique(forecast_data$Quarter_Year))),
    rep(operational_capacity, length(unique(forecast_data$Quarter_Year)))
  ),
  Type = c(
    rep("Absolute Maximum Capacity", length(unique(forecast_data$Quarter_Year))),
    rep("Emergency Capacity", length(unique(forecast_data$Quarter_Year))),
    rep("Operational Capacity", length(unique(forecast_data$Quarter_Year)))
  )
)

capacity_reached <- list(
  absolute_max = NA,
  emergency = NA,
  operational = NA
)

for (i in 1:length(horizon_dates)) {
  date <- horizon_dates[i]
  projected_value <- forecast_data$Forecasted[forecast_data$Quarter_Year == date]

  if (!is.na(projected_value)) {
    if (is.na(capacity_reached$absolute_max) && projected_value > absolute_max_capacity) {
      capacity_reached$absolute_max <- date
    }
    if (is.na(capacity_reached$emergency) && projected_value > emergency_capacity) {
      capacity_reached$emergency <- date
    }
    if (is.na(capacity_reached$operational) && projected_value > operational_capacity) {
      capacity_reached$operational <- date
    }
  }
}

p <- plot_ly(
  hoverlabel = list(namelength = -1)
) %>%
  add_trace(
    data = forecast_data[!is.na(forecast_data$Actual),],
    x = ~Quarter_Year,
    y = ~Actual,
    type = 'scatter',
    mode = 'lines',
    name = "Historical",
    line = list(color = color_actual),
    hoverinfo = "text",
    text = ~paste0(
      format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
      "<br>Population: ", formatC(Actual, format="d", big.mark=",")
    )
  )

forecast_part <- forecast_data[forecast_data$Quarter_Year %in% horizon_dates,]
p <- p %>%
  add_trace(
    data = forecast_part,
    x = ~Quarter_Year,
    y = ~Forecasted,
    type = 'scatter',
    mode = 'lines',
    name = "Projection",
    line = list(color = "#A47217", dash = "dash"),
    hoverinfo = "text",
    text = ~paste0(
      format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
      "<br>Projected: ", formatC(Forecasted, format="d", big.mark=",")
    )
  )

p <- p %>%
  add_trace(
    data = capacity_df[capacity_df$Type == "Absolute Maximum Capacity",],
    x = ~Quarter_Year, y = ~Capacity,
    type = 'scatter', mode = 'lines',
    name = "Absolute Maximum Capacity",
    line = list(color = cap_col_max, dash = "dot", width = 2),
    hoverinfo = "text",
    text = ~paste0(
      format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
      "<br>Absolute Maximum Capacity: ", formatC(Capacity, format="d", big.mark=",")
    )
  ) %>%
  add_trace(
    data = capacity_df[capacity_df$Type == "Emergency Capacity",],
    x = ~Quarter_Year, y = ~Capacity,
    type = 'scatter', mode = 'lines',
    name = "Emergency Capacity",
    line = list(color = cap_col_emerg, dash = "dot", width = 2),
    hoverinfo = "text",
    text = ~paste0(
      format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
      "<br>Emergency Capacity: ", formatC(Capacity, format="d", big.mark=",")
    )
  ) %>%
  add_trace(
    data = capacity_df[capacity_df$Type == "Operational Capacity",],
    x = ~Quarter_Year, y = ~Capacity,
    type = 'scatter', mode = 'lines',
    name = "Operational Capacity",
    line = list(color = cap_col_oper, dash = "dot", width = 2),
    hoverinfo = "text",
    text = ~paste0(
      format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
      "<br>Operational Capacity: ", formatC(Capacity, format="d", big.mark=",")
    )
  )

for (capacity_type in names(capacity_reached)) {
  date <- capacity_reached[[capacity_type]]
  if (!is.na(date)) {
    projected_value <- forecast_data$Forecasted[forecast_data$Quarter_Year == date]

    year <- format(date, "%Y")
    quarter <- paste0("Q", ceiling(as.numeric(format(date, "%m"))/3))
    quarter_label <- paste0(year, " ", quarter)

    capacity_name <- switch(capacity_type,
                            "absolute_max" = "Absolute Maximum Capacity",
                            "emergency"    = "Emergency Capacity",
                            "operational"  = "Operational Capacity")

    annotation_color <- switch(capacity_type,
                               "absolute_max" = cap_col_max,
                               "emergency"    = cap_col_emerg,
                               "operational"  = cap_col_oper)

    p <- p %>%
      add_annotations(
        x = date,
        y = projected_value,
        text = paste(quarter_label),
        showarrow = TRUE,
        arrowhead = 2,
        arrowsize = 1,
        arrowwidth = 1.5,
        arrowcolor = annotation_color,
        ax = switch(capacity_type,
                    "absolute_max" =  70,
                    "emergency"    = -80,
                    "operational"  = -80),
        ay = switch(capacity_type,
                    "absolute_max" = -70,
                    "emergency"    = -35,
                    "operational"  =  45),
        font = list(color = annotation_color, size = 11,
                    family = "UT Industry, serif"),
        bgcolor = "rgba(255, 255, 255, 0.88)",
        bordercolor = annotation_color,
        borderwidth = 1,
        borderpad = 4,
        align = "center"
      )
  }
}

all_quarters <- unique(forecast_data$Quarter_Year)

y_top <- max(c(absolute_max_capacity, forecast_data$Forecasted, forecast_data$Actual),
             na.rm = TRUE) * 1.03

p %>% layout(
  title = list(
    text = title,
    font = list(family = "UT Industry, serif", size = 18, color = "#071D49"),
    y = 0.96,
    yanchor = "top"
  ),
  margin = list(t = 60, b = 50, l = 70, r = 30),
  font = list(family = "UT Industry, serif", size = 11),
  showlegend = TRUE,
  shapes = list(list(
    type = "rect", xref = "paper", yref = "y",
    x0 = 0, x1 = 1, y0 = operational_capacity, y1 = y_top,
    fillcolor = "rgba(155, 27, 21, 0.07)",
    line = list(width = 0), layer = "below"
  )),
  xaxis = list(
    title = "Year-Quarter",
    type = "date",
    tickmode = "array",
    tickvals = all_quarters[seq(1, length(all_quarters), by = 8)],
    ticktext = format(all_quarters[seq(1, length(all_quarters), by = 8)], "%Y"),
    tickangle = 0
  ),
  yaxis = list(
    title = "Prison Population",
    tickformat = ",",
    range = c(5000, 7600)      # shared with the total chart for comparability
  ),
  hovermode = "closest"
)
```

## Female Population {height="420px"}

```{r}
#| label: female-capacity-analysis

female_data <- get_data_for_population("female")
forecast_data <- female_data$forecast_data
color_actual <- female_data$color
title <- paste(female_data$title, "vs. Capacity")

absolute_max_capacity <- 632   # Female-specific absolute maximum capacity
emergency_capacity <- 619      # VERIFY against UDC source
operational_capacity <- 610    # VERIFY against UDC source

actual_data <- forecast_data[!is.na(forecast_data$Actual),]
max_date <- max(actual_data$Quarter_Year)

horizon_dates <- seq(max_date + 1, by = 1, length.out = projection_horizon)

capacity_df <- data.frame(
  Quarter_Year = rep(unique(forecast_data$Quarter_Year), 3),
  Capacity = c(
    rep(absolute_max_capacity, length(unique(forecast_data$Quarter_Year))),
    rep(emergency_capacity, length(unique(forecast_data$Quarter_Year))),
    rep(operational_capacity, length(unique(forecast_data$Quarter_Year)))
  ),
  Type = c(
    rep("Absolute Maximum Capacity", length(unique(forecast_data$Quarter_Year))),
    rep("Emergency Capacity", length(unique(forecast_data$Quarter_Year))),
    rep("Operational Capacity", length(unique(forecast_data$Quarter_Year)))
  )
)

capacity_reached <- list(
  absolute_max = NA,
  emergency = NA,
  operational = NA
)

for (i in 1:length(horizon_dates)) {
  date <- horizon_dates[i]
  projected_value <- forecast_data$Forecasted[forecast_data$Quarter_Year == date]

  if (!is.na(projected_value)) {
    if (is.na(capacity_reached$absolute_max) && projected_value > absolute_max_capacity) {
      capacity_reached$absolute_max <- date
    }
    if (is.na(capacity_reached$emergency) && projected_value > emergency_capacity) {
      capacity_reached$emergency <- date
    }
    if (is.na(capacity_reached$operational) && projected_value > operational_capacity) {
      capacity_reached$operational <- date
    }
  }
}

p <- plot_ly(
  hoverlabel = list(namelength = -1)
) %>%
  add_trace(
    data = forecast_data[!is.na(forecast_data$Actual),],
    x = ~Quarter_Year,
    y = ~Actual,
    type = 'scatter',
    mode = 'lines',
    name = "Historical",
    line = list(color = color_actual),
    hoverinfo = "text",
    text = ~paste0(
      format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
      "<br>Population: ", formatC(Actual, format="d", big.mark=",")
    )
  )

forecast_part <- forecast_data[forecast_data$Quarter_Year %in% horizon_dates,]
p <- p %>%
  add_trace(
    data = forecast_part,
    x = ~Quarter_Year,
    y = ~Forecasted,
    type = 'scatter',
    mode = 'lines',
    name = "Projection",
    line = list(color = "#A47217", dash = "dash"),
    hoverinfo = "text",
    text = ~paste0(
      format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
      "<br>Projected: ", formatC(Forecasted, format="d", big.mark=",")
    )
  )

p <- p %>%
  add_trace(
    data = capacity_df[capacity_df$Type == "Absolute Maximum Capacity",],
    x = ~Quarter_Year, y = ~Capacity,
    type = 'scatter', mode = 'lines',
    name = "Absolute Maximum Capacity",
    line = list(color = cap_col_max, dash = "dot", width = 2),
    hoverinfo = "text",
    text = ~paste0(
      format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
      "<br>Absolute Maximum Capacity: ", formatC(Capacity, format="d", big.mark=",")
    )
  ) %>%
  add_trace(
    data = capacity_df[capacity_df$Type == "Emergency Capacity",],
    x = ~Quarter_Year, y = ~Capacity,
    type = 'scatter', mode = 'lines',
    name = "Emergency Capacity",
    line = list(color = cap_col_emerg, dash = "dot", width = 2),
    hoverinfo = "text",
    text = ~paste0(
      format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
      "<br>Emergency Capacity: ", formatC(Capacity, format="d", big.mark=",")
    )
  ) %>%
  add_trace(
    data = capacity_df[capacity_df$Type == "Operational Capacity",],
    x = ~Quarter_Year, y = ~Capacity,
    type = 'scatter', mode = 'lines',
    name = "Operational Capacity",
    line = list(color = cap_col_oper, dash = "dot", width = 2),
    hoverinfo = "text",
    text = ~paste0(
      format(Quarter_Year, "%Y Q"), ceiling(as.numeric(format(Quarter_Year, "%m"))/3),
      "<br>Operational Capacity: ", formatC(Capacity, format="d", big.mark=",")
    )
  )

for (capacity_type in names(capacity_reached)) {
  date <- capacity_reached[[capacity_type]]
  if (!is.na(date)) {
    projected_value <- forecast_data$Forecasted[forecast_data$Quarter_Year == date]

    year <- format(date, "%Y")
    quarter <- paste0("Q", ceiling(as.numeric(format(date, "%m"))/3))
    quarter_label <- paste0(year, " ", quarter)

    capacity_name <- switch(capacity_type,
                            "absolute_max" = "Absolute Maximum Capacity",
                            "emergency"    = "Emergency Capacity",
                            "operational"  = "Operational Capacity")

    annotation_color <- switch(capacity_type,
                               "absolute_max" = cap_col_max,
                               "emergency"    = cap_col_emerg,
                               "operational"  = cap_col_oper)

    p <- p %>%
      add_annotations(
        x = date,
        y = projected_value,
        text = paste(quarter_label),
        showarrow = TRUE,
        arrowhead = 2,
        arrowsize = 1,
        arrowwidth = 1.5,
        arrowcolor = annotation_color,
        ax = switch(capacity_type,
                    "absolute_max" =  70,
                    "emergency"    = -80,
                    "operational"  = -80),
        ay = switch(capacity_type,
                    "absolute_max" = -70,
                    "emergency"    = -35,
                    "operational"  =  45),
        font = list(color = annotation_color, size = 11,
                    family = "UT Industry, serif"),
        bgcolor = "rgba(255, 255, 255, 0.88)",
        bordercolor = annotation_color,
        borderwidth = 1,
        borderpad = 4,
        align = "center"
      )
  }
}

all_quarters <- unique(forecast_data$Quarter_Year)

y_top <- max(c(absolute_max_capacity, forecast_data$Forecasted, forecast_data$Actual),
             na.rm = TRUE) * 1.03

p %>% layout(
  title = list(
    text = title,
    font = list(family = "UT Industry, serif", size = 18, color = "#071D49"),
    y = 0.96,
    yanchor = "top"
  ),
  margin = list(t = 60, b = 50, l = 70, r = 30),
  font = list(family = "UT Industry, serif", size = 11),
  showlegend = TRUE,
  shapes = list(list(
    type = "rect", xref = "paper", yref = "y",
    x0 = 0, x1 = 1, y0 = operational_capacity, y1 = y_top,
    fillcolor = "rgba(155, 27, 21, 0.07)",
    line = list(width = 0), layer = "below"
  )),
  xaxis = list(
    title = "Year-Quarter",
    type = "date",
    tickmode = "array",
    tickvals = all_quarters[seq(1, length(all_quarters), by = 8)],
    ticktext = format(all_quarters[seq(1, length(all_quarters), by = 8)], "%Y"),
    tickangle = 0
  ),
  yaxis = list(
    title = "Prison Population",
    tickformat = ",",
    range = c(350, 700)        # own scale — female never approaches the male range
  ),
  hovermode = "closest"
)
```


# 📝 Methodology

## Row

```{r}
#| label: pdf-embed
#| echo: false

# Title + PDF emitted as ONE block so the card doesn't space them apart.
htmltools::tagList(
  htmltools::tags$h3(
    "Prison Population Projection Methodology",
    style = "color:#071D49; font-family:'UT Industry', serif; font-weight:700; margin:0 0 0.85rem 0;"
  ),
  htmltools::tags$iframe(
    src = "2026-UPPP-Methodology.pdf",
    width = "100%",
    height = "100%",
    style = "border:1px solid #ddd; display:block; min-height:72vh;"
  )
)
```


